# Starting With Data

R has built in functions to bring in data like `read.csv()`, but because ADaM datasets are commonly stored as `.sas7bdat` files or `.xpt` files, we will need to install the package `haven`. This package enables R to read and write SAS and XPT formats. To install a package we call the function `install.packages` (we only need to do this once), then to use that package we must call it using `library`. 

```{r, results = "hide"}
# install.packages("haven")
library(haven)
```

Let's start with out ADSL file which we placed in `raw_data`, and save it as an object. 

```{r}
ADSL <- read_xpt("data_raw/adsl.xpt")
```

This statement doesn't produce any output because assignments don't display anything. If we want to check that our data has loaded we can see the contents by typing the name of the data... but that's a lot of output!

Instead we'll check the top (the first 6 lines) of our datasets using the function `head()`

```{r}
head(ADSL)
```

Try also `View(ADSL)`

## What is a data frame? 

Data frames are the de facto data structure for most tabular data, and what we use for statistics and plotting.

A data frame can be created by hand, but most commonly they are generated after importing spreadsheets from your hard drive (or the web).

A data frame is the representation of data in the format of a table where the columns are vectors that all have the same length. Because columns are vectors, each column must contain a single type of data (e.g., characters, integers, factors). For example, here is a figure depicting a data frame comprising a numeric, a character, and a logical vector.

We can see this when inspecting the structure of a data frame with the function `str()`:

```{r}
str(ADSL)
```

We already saw how the functions `head()` and `str()` can be useful to check the content and the structure of a data frame. Here is a non-exhaustive list of functions to get a sense of the content/structure of the data. Let’s try them out!

* Size:
    * dim(ADSL) - returns a vector with the number of rows in the first element, and the number of columns as the second element (the dimensions of the object)
    * nrow(ADSL) - returns the number of rows
    * ncol(ADSL) - returns the number of columns
    
* Content:
    * head(ADSL) - shows the first 6 rows
    * tail(ADSL) - shows the last 6 rows
    
* Names:
    * names(ADSL) - returns the column names (synonym of colnames() for data.frame objects)
    * rownames(ADSL) - returns the row names
    
* Summary:
    * str(ADSL) - structure of the object and information about the class, length and content of each column
    * summary(ADSL) - summary statistics for each column
    
Note: most of these functions are “generic”, they can be used on other types of objects besides `data.frame`.

## Exercise 

Based on the output of `str(ADSL)`, can you answer the following questions?

* What is the class of the object ADSL?
* How many rows and how many columns are in this object?

## Indexing and Subsetting Data Frames

Our data frame has rows and columns (it has 2 dimensions), if we want to extract some specific data from it, we need to specify the “coordinates” we want from it. Row numbers come first, followed by column numbers. However, note that different ways of specifying these coordinates lead to results with different classes.

```{r}
# first element in the first column of the data frame (as a vector)
ADSL[1, 1]   
```

```{r}
# first element in the 6th column (as a vector)
ADSL[1, 6]   
```

```{r}
# first column of the data frame (as a vector)
ADSL[, 1] 
```

```{r}
# first column of the data frame (as a data.frame)
ADSL[1]    
```

```{r}
# first three elements in the 7th column (as a vector)
ADSL[1:3, 7] 
```

```{r}
# the 3rd row of the data frame (as a data.frame)
ADSL[3, ]
```

```{r}
# equivalent to head_ADSL <- head(ADSL)
head_ADSL <- ADSL[1:6, ] 
```

`:` is a special function that creates numeric vectors of integers in increasing or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain indices of a data frame using the `-` sign:

```{r}
ADSL[, -1]          # The whole data frame, except the first column
```

```{r}
ADSL[-c(7:254), ]   # Equivalent to head(surveys)
```

Data frames can be subset by calling indices (as shown previously), but also by calling their column names directly:

```{r}
ADSL["ARM"]       # Result is a data.frame
```

```{r}
ADSL[, "ARM"]     # Result is a vector
```

```{r}
ADSL[["ARM"]]     # Result is a vector
```

```{r}
ADSL$ARM          # Result is a vector
```

## Exercise

1. Ce ate a `data.frame` (`ADSL_50`) containing only the data in row 50 of the `ADSL` dataset.

2. Notice how nrow() gave you the number of rows in a data.frame?
  * Use that number to pull out just that last row in the data frame.
  * Compare that with what you see as the last row using `tail()` to make sure it’s meeting expectations.
  * Pull out that last row using `nrow()` instead of the row number.
  * Create a new data frame (`ADSL_last`) from that last row.
  * Use `nrow()` to extract the row that is in the middle of the data frame. Store the content of this row in an object named `ADSL_middle`.
  
3. Combine `nrow()` with the `-` notation above to reproduce the behavior of head(ADSL), keeping just the first through 6th rows of the ADSL dataset.

## Factors
